---
title: "Chapter 5: Deviating from the Average"
author: "Wendy Graham"
date: "2025-07-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load Libraries
```{r}
install.packages("modeest")

library(MASS)
library(ggplot2)
library(modeest)
```

## Measuring Variation

To develop the variance formula for you and show you how it works, I use symbols to show all this:
$$
(X - \overline{X})
$$

$$(X - \overline{X})^2
$$

To calculate variance, you square each deviation, add them up, and find the average of the squared deviations. If N represents the amount of squared deviations you have (in this example, five), the formula for calculating the variance is

$$
\sigma^2 = \frac{\sum (X - \overline{X})^2}{N}
$$
The ∑ character is the uppercase Greek letter sigma, and it means “the sum of.”

## Sample Variance
Sample variance (as an estimate of the population variance) is
$$
s^2 = \frac{\sum (X - \overline{X})^2}{n - 1}
$$



```{r variance}
#sample variance
heights <- c(50, 47, 52, 46, 45)
var(heights)

#population variance
var(heights)*(length(heights)-1)/length(heights)

#defining population variance
var.p = function(x){var(x)*(length(x)-1)/length(x)}

```

## Standard Deviation

After you calculate the variance of a set of numbers, you have a value whose units are different from your original measurements. For example, if your original measurements are in inches, their variance is in square inches. This is because you square the deviations before you average them. So the variance in the 5-score population in the preceding example is 6.8 square inches.

It might be hard to grasp what that statement means. Often, it’s more intuitive if the variation statistic is in the same units as the original measurements. It’s easy to turn variance into that kind of statistic. All you have to do is take the square root of the variance.

Like the variance, this square root is so important that it has a special name: standard deviation.

$$
\sigma^2 = \frac{\sum (X - \overline{X})^2}{N}
$$

$$
s^2 = \frac{\sum (X - \overline{X})^2}{n - 1}
$$

```{r standard deviation}
#sample standard deviation
sd(heights)

#population standard deviation
sd(heights)*(sqrt((length(heights)-1)/length(heights)))

#define a population standard deviation
sd.p=function(x){sd(x)*sqrt((length(x)-1)/length(x))}
```


##Conditions, Conditions, Conditions …


```{r with conditions}
print("Variance of horsepower for USA cars")
with(Cars93, var(Horsepower[Origin == "USA"]))

print("How Many Cars in this group?")
with(Cars93, length(Horsepower[Origin == "USA"]))

print("Variance of horsepower for non-USA cars")
with(Cars93, var(Horsepower[Origin == "non-USA"]))

print("How Many Cars in this group?")
with(Cars93, length(Horsepower[Origin == "non-USA"]))
```


## Experimental code

```{r sd with conditions}
print("Horsepower standard deviation for USA cars")
with(Cars93, sd(Horsepower[Origin == "USA"]))


print("Horsepower standard deviation for non-USA cars")
with(Cars93, sd(Horsepower[Origin == "non-USA"]))

```

